---
title: "Visualizations using SCE outage data for CPA jurisdictions"
author: "David Segan"
date: "4/27/2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, message = FALSE)
```

```{r}
# --------------
# attach packages
# --------------
library(tidyverse)
library(readxl)
library(janitor)
library(paletteer)
library(dplyr)
library(scales)
library(extrafont)
library(gfonts)
library(formattable)
get_all_fonts()

```

```{r}
# ---------
# Read in data
# ---------

hist_outages <- read_excel("SCE_Reliability_Metrics.xlsx") %>% 
  janitor::clean_names() %>% 
  mutate(jurisdiction = str_remove_all(jurisdiction, "Unincorporated")) %>% 
  mutate(total_outage_hrs = total_customers * sce_average_saidi_hours) %>% 
  mutate(natdis_psps_outages = percent_saifi_psps_weather_fire_earthquake2*sce_average_saifi_interruptions)

# --------
# Check value types
# --------

# class(hist_outages$year) # numeric
# class(hist_outages$percent_saidi_psps) # numeric
```

On average, what jurisdictions have had the most time per customer without power from 2016 through 2019? 
```{r}
# data frame
avg_outages <- hist_outages %>% 
  group_by(jurisdiction, county) %>% 
  summarize(mean_saidi_hrs = mean(sce_average_saidi_hours)) %>% 
  arrange(-mean_saidi_hrs)

# graph
avg_outages_graph <- ggplot(data = avg_outages,
                            aes(x = mean_saidi_hrs,
                                y = fct_reorder(jurisdiction,mean_saidi_hrs),
                                fill = county)) +
  geom_col() + 
  theme_minimal() +
  scale_fill_manual(values = c("orange", "cornflowerblue")) +
  scale_x_continuous(lim = c(0,60), expand = c(0,0)) +
  labs(x = "Hours",
       y = "Jurisdiction",
       title = "Average Annual Outage Time \nPer Customer (2016-2019)",
       fill = "County")
  
avg_outages_graph

# ggsave("avg_annual_outage_duration_barchart.jpg")

# Note outlier: In 2018, residents of Malibu served by SCE had ~211 hours without grid electricity
```

What jurisdiction had the most overall outage hours? This could be considered total damage (as opposed to per person in previous chart), as it takes into account population.
```{r}
# data frame
tot_outages <- hist_outages %>% 
  group_by(jurisdiction, county) %>% 
  # Total outage hours is equal to total customers x average outage time per customer (SAIDI)
  # total_outage_hrs was calculated outside of R
  summarize(total_outage_hrs = mean(total_outage_hrs)) %>% 
  arrange(-total_outage_hrs)


# graph
tot_outages_graph <- ggplot(data = tot_outages,
                            aes(x = total_outage_hrs,
                                y = fct_reorder(jurisdiction,total_outage_hrs),
                                fill = county)) +
  geom_col() + 
  theme_minimal() +
  scale_fill_manual(values = c("orange", "cornflowerblue")) +
  scale_x_continuous(lim = c(0,900000), expand = c(0,0), labels = comma) +
  labs(x = "Hours",
       y = "Jurisdiction",
       title = "Total Outage Time \n(Averaged across 2016-2019)",
       fill = "County")

tot_outages_graph

# ggsave("tot_outage_time_barchart.jpg")
```

What are the causes of outages in CPA service territory? 
```{r}
# data frame
outage_causes <- hist_outages %>% 
  # add columns that sum hrs due to specific causes
  mutate(hrs_psps = total_outage_hrs * percent_saidi_psps) %>% 
  mutate(hrs_weather_fire_quake = total_outage_hrs * percent_saidi_weather_fire_earthquake) %>% 
  mutate(hrs_veg_animal = total_outage_hrs * percent_saidi_vegetation_animal) %>% 
  mutate(hrs_other = total_outage_hrs * percent_saidi_other) %>% 
  mutate(hrs_equipment = total_outage_hrs * percent_saidi_equipment_failure) %>% 
  mutate(hrs_3rdparty = total_outage_hrs * percent_saidi_3rd_party) %>% 
  # select only columns needed
  select(jurisdiction, total_customers, year, hrs_psps,hrs_weather_fire_quake, hrs_veg_animal, hrs_other, hrs_equipment, hrs_3rdparty) %>% 
  # group and sum by year
  group_by(year) %>% 
  summarize(hrs_psps = (sum(hrs_psps, na.rm = TRUE)/sum(total_customers)),
            hrs_weather_fire_quake = (sum(hrs_weather_fire_quake, na.rm = TRUE)/sum(total_customers)),
            hrs_veg_animal = (sum(hrs_veg_animal, na.rm = TRUE)/sum(total_customers)),
            hrs_other = (sum(hrs_other, na.rm = TRUE)/sum(total_customers)),
            hrs_equipment = (sum(hrs_equipment, na.rm = TRUE)/sum(total_customers)),
            hrs_3rdparty = (sum(hrs_3rdparty, na.rm = TRUE)/sum(total_customers))) %>% 
  # pivot longer to have column for cause and column for hours
  pivot_longer('hrs_psps':'hrs_3rdparty',
               names_to = "cause",
               values_to = "hours")

outage_causes_graph <- ggplot(data = outage_causes,
                              aes(x = year,
                                  y = hours)) +
  geom_col(aes(fill = cause)) + 
  theme_minimal() +
  # scale_fill_paletteer_d("ggsci::default_jama") +
  labs(x = "Year",
       y = "Hours per Customer",
       title = "The causes of outages in CPA service territory",
       fill = "Cause") +
  # manually entered hex codes of ggsci default jama using this webpage: https://thefeeney.netlify.app/post/color-palettes-in-r/
  scale_fill_manual(values = c("#374E55FF", "#80796BFF", "#00A1D5FF", "#B24745FF", "#79AF97FF", "#DF8F44FF"), labels = c("3rd Party", "Equipment Failure", "Other", "PSPS", "Vegetation or Animal", "Weather, Fire, or Earthquake")) +
  theme(legend.title = element_text(size = 16),
        legend.text = element_text(size = 15),
        axis.title = element_text(size = 16),
        title = element_text(size = 18),
        axis.text = element_text(size = 15))

outage_causes_graph

# ggsave("outage_causes_stacked_bar_chart.jpg")
```

On average, how long will it take to restore power? 
```{r}
# data frame

# add caidi values # "CAIDI" = (SAIDI/SAIFI) 
caidi <- hist_outages %>% 
  mutate(caidi_minutes = sce_average_saidi_minutes/sce_average_saifi_interruptions) %>% 
  mutate(caidi_hrs = sce_average_saidi_hours/sce_average_saifi_interruptions) %>% 
  group_by(jurisdiction, county) %>% 
  summarize(mean_caidi_hrs = mean(caidi_hrs)) %>% 
  arrange(-mean_caidi_hrs)

# graph
caidi_graph <- ggplot(data = caidi,
                            aes(x = mean_caidi_hrs,
                                y = fct_reorder(jurisdiction, mean_caidi_hrs),
                                fill = county)) +
  geom_col() + 
  theme_minimal() +
  scale_fill_manual(values = c("orange", "cornflowerblue")) +
  scale_x_continuous(lim = c(0,13), expand = c(0,0)) +
  labs(x = "Hours",
       y = "Jurisdiction",
       title = "Average Time to Restore Power (2016-2019)",
       fill = "County") +
  theme_minimal() +
  theme(
    plot.subtitle = element_text(hjust = 0),
    legend.background = element_rect(fill="white", linetype="solid", colour = "white"),
    legend.position = c(.85,.26),
    # plot.margin=unit(c(.5,1.1,.5,.5),"cm"),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 15),
        axis.title = element_text(size = 16),
        title = element_text(size = 18),
        axis.text = element_text(size = 13))

caidi_graph

# ggsave("avg_caidi_barchart.jpg")
```

On average, how many outages occur?
```{r}
# data frame
# SAIFI = frequency
saifi <- hist_outages %>% 
  group_by(jurisdiction, county) %>% 
  summarize(mean_saifi = mean(sce_average_saifi_interruptions)) %>% 
  arrange(-mean_saifi)

# graph
avg_saifi_graph <- ggplot(data = saifi,
                            aes(x = mean_saifi,
                                y = fct_reorder(jurisdiction,mean_saifi),
                                fill = county)) +
  geom_col() + 
  theme_minimal() +
  scale_fill_manual(values = c("orange", "cornflowerblue")) +
  scale_x_continuous(lim = c(0,4), expand = c(0,0)) +
  labs(x = "Outages",
       y = "Jurisdiction",
       title = "Average Number of Outages Per Year",
       subtitle = "Average across 2016-2019",
       fill = "County")

avg_saifi_graph

# ggsave("avg_saifi_barchart.jpg")
```

```{r}
# Have more frequent outages caused by PSPS events, Weather, Fire, or Earthquake been associated with longer outage durations?

# Ideally, we would have data on specific PSPS/Weather/Fire/Earthquake outages.
# This data was not used for any correlation analysis or included in the final report

#graph
# natdis_scatter_all_yrs <- ggplot(data = hist_outages,
#                           aes(x = natdis_psps_outages,
#                               y = caidi_hours,
#                               color = year)) +
#   geom_point() +

#   theme_minimal() +
#   scale_color_paletteer_d("ggthemes::colorblind") +
#   scale_x_continuous(lim = c(0,2), expand = c(0,0)) +
#   geom_smooth(method = "lm") +
#   labs(x = "Average # of Outages Caused by PSPS, Weather, Fire, or Earthquake",
#        y = "Customer Average Outage Duration (CAIDI)",
#        title = "Natural disaster caused outages affect duration",
#        subtitle = "2016-2019",
#        color = "Year")
# 
# natdis_scatter_all_yrs

```

How has outage length changed over time across all of SCE service territory?
```{r}
# Data is available over a longer time period for all of SCE territory in a different data set. This is as opposed to the jurisdiction specific data that has been looked at in prior graphs.

# read in data
eia_ca_reliability <- read_excel("EIA_Utility_Reliability_Data.xlsx", skip = 1) %>% 
  clean_names() %>% 
  rename(year = data_year) %>% 
  filter(state == "CA") %>% 
  # change all numbers from character to numeric
  mutate_at(vars(matches("med")), funs(as.numeric)) %>% 
  mutate(saidi_hrs = saidi_with_med_6/60)

# filter for just SCE
eia_sce_reliability <- eia_ca_reliability %>% 
  filter(utility_name == "Southern California Edison Co") 

#graph
sce_outage_plot <- ggplot(data = eia_sce_reliability,
                          aes(x = year,
                              y = saidi_hrs)) +
  geom_line() + 
  theme_minimal() +
  theme(
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 14),
    plot.margin=unit(c(.5,1.1,.5,.5),"cm"),
    axis.title = element_text(size = 15),
    title = element_text(size = 18),
    axis.text = element_text(size = 13)) +
  scale_x_continuous(expand = c(0,.1)) +
  scale_y_continuous(lim = c(1.5,3)) +
  labs(x = "Year",
       y = "Average Customer Outage \nDuration (hours/year)",
       title = "Upward Trend of Outage Lengths in SCE Service Territory (2013-2019)")

sce_outage_plot

# ggsave("sce_outage_plot.jpg")
```

What is the value of backup power for each jurisdiction based on average historical outage trends?

```{r}
# df that contains data for cost of outage and value of resilience equations
caidi_vor <- hist_outages %>% 
  group_by(jurisdiction, county) %>% 
  summarize(mean_caidi_hrs = mean(caidi_hours), mean_saidi_hrs = mean(sce_average_saidi_hours), mean_saifi = mean(sce_average_saifi_interruptions),
            mean_pct_natdis_psps = mean(percent_saidi_psps_weather_fire_earthquake)) %>% 
  arrange(-mean_caidi_hrs)

# data frame
vor_equation <- caidi_vor %>% 
  # discount rate so future year value can be discounted to present day (to compare to upfront cost of system and islanding)
  mutate(discount_rate = .05) %>% 
  #liftime of system
  mutate(system_life = 10) %>% 
  # cost_unserved_kWh will be synonymous with VoLL ($/kWh). The term Voll is used in the final report
  # see final report for explanation on the benefit transfer method used to determine these values
  # --------------------
  # if possible, these cost_unserved_kWh values should be changed based on specific critical facility through the stated preference method (survey) or analysis of the value of the service provided
  # --------------------
  mutate(cost_unserved_kwh_1hr = 21.8) %>% 
  mutate(cost_unserved_kwh_4hr = 12.1) %>% 
  mutate(cost_unserved_kwh_8hr = 12.9) %>% 
  mutate(cost_unserved_kwh_16hr = 12.7) %>% 
  # interpolation
  mutate(cost_unserved_kwh_caidi = if(mean_caidi_hrs < 4) {
    (cost_unserved_kwh_1hr+((mean_caidi_hrs-1)/(4-1))*(cost_unserved_kwh_4hr-cost_unserved_kwh_1hr))
  } else if (mean_caidi_hrs >= 4 & mean_caidi_hrs < 8) {
    (cost_unserved_kwh_4hr+((mean_caidi_hrs-4)/(8-4))*(cost_unserved_kwh_8hr-cost_unserved_kwh_4hr))
  } else if (mean_caidi_hrs >= 8 & mean_caidi_hrs < 16) {
    (cost_unserved_kwh_8hr+((mean_caidi_hrs-8)/(16-8))*(cost_unserved_kwh_16hr-cost_unserved_kwh_8hr))
  } else {
    cost_unserved_kwh_16hr
  }) %>% 
  # --------------------
  # avg_load_kw should be changed based on specific critical facility
  # --------------------
  mutate(avg_load_kw = 815) %>% 
  # --------------------
  # avg_critical_load_kw should be changed based on specific critical facility
  # --------------------
  mutate(avg_critical_load_kw = 50) %>% 
  mutate(cost_outage_per_hour = avg_load_kw*cost_unserved_kwh_caidi) %>% 
  mutate(cost_outage = avg_load_kw*cost_unserved_kwh_caidi*mean_caidi_hrs) %>% 
  # --------------------
  # pv_storage_backup_hrs should be changed based on specific pv+s system and critical_load_kw
  # --------------------
  mutate(pv_storage_backup_hrs = 4) %>% 
  mutate(effective_backup_hrs = if((mean_caidi_hrs-pv_storage_backup_hrs) <0) {
    mean_caidi_hrs
  } else {
    pv_storage_backup_hrs
  }) %>% 
  mutate(value_resilience = cost_unserved_kwh_caidi*avg_critical_load_kw*effective_backup_hrs) %>% 
  mutate(value_resilience_thousands = value_resilience/1000) %>% 
  mutate(lifetime_value_resilience = (value_resilience*mean_saifi)*((1-(1+discount_rate)^-system_life)/discount_rate)) %>% 
  mutate(mean_pct_natdis_psps = formattable::percent(mean_pct_natdis_psps))

# graph
vor_graph <- ggplot(data = vor_equation, 
                                          aes(x = lifetime_value_resilience,
                                              y = fct_reorder(jurisdiction,lifetime_value_resilience),
                                              fill = mean_pct_natdis_psps)) +
  geom_col() + 
  scale_fill_paletteer_c("ggthemes::Red", labels = percent) +
  scale_x_continuous(lim = c(0,80000), expand = c(0,0), labels = dollar_format()) +
  labs(x = "Value of Resilience",
       y = "Jurisdiction",
       title = "The value of resilience varies by community",
       subtitle = "Modeled scenario over 10 years with 50kw of critical load that is sustained \nfor up to 4 hours each outage. The cost of unserved kWh ranges between \n$21.8 and $12.7 depending on outage length, and the discount rate is 5%",
       fill = "Average % contribution to \noutage time from PSPS \nevents & natural disasters") +
  theme_minimal() +
  theme(
    legend.background = element_rect(fill="white", linetype="solid", colour = "white"),
    legend.position = c(.81,.31),
    plot.margin=unit(c(.5,1.1,.5,.5),"cm"),
    legend.title = element_text(size = 18),
    legend.text = element_text(size = 18),
    axis.title = element_text(size = 15),
    plot.subtitle = element_text(size = 13, hjust = 0),
    title = element_text(size = 18),
    axis.text = element_text(size = 13))

vor_graph

# ggsave("vor_graph\.jpg", width = 18.5, height = 9.5, units = "in")
```


